Dependent Object Types

### 2012-08-05 article.tex
update counterex to latest calculus
formatting
add numbered subsections throughout
a bit of discussion

\subsection*{Why No Inheritance?}

@@

\subsection*{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
%introduce a variable which is already bound in the base-set.
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

\subsection*{Syntax}

@@

\subsection*{Reduction rules}

@@

\subsection*{Type assignment rules}

@@

\subsection*{Membership}

@@

\subsection*{Expansion}

@@

\subsection*{Subtyping}

@@

\subsection*{Declaration Subsumption}

@@

\subsection*{Well-formedness}

@@

\begin{align*}
&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
\mlapp {\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \abs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\abs a {y.A} {\app {(\abs x \Top x)} a}}}}{
\ \app {(\abs x \Top x)} z
}}}{\ v}}
\end{align*}

@@

>>>>

\subsection{Why No Inheritance?}

@@

\subsection{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
%introduce a variable which is already bound in the base-set.
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

\subsection{Syntax}

@@

\subsection{Reduction rules}

@@

\subsection{Type assignment rules}

@@

\subsection{Membership}

@@

\subsection{Expansion}

@@

\subsection{Subtyping}

@@

\subsection{Declaration Subsumption}

@@

\subsection{Well-formedness}

@@

\begin{align*}
&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
\mlapp {(\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
\mlnew z {\Top \mlrefine z {\\&\gap\gap\gap
\mdecl l {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} \Top}\mlldefs{\\&\gap\gap\gap
l(y) = \fun a {y.A} \Top a}}{
\gap \cast {\top} z
}})}{\ v}}
\end{align*}

@@

\subsection{Variants of the DOT Calculus}\label{dot-variants}

\subsubsection{Why limit the calculus to $\wfe$-types?}

Currently, the proof of type-safety via logical relations
fundamentally relies on types having an expansion. However, this was
not our original motivation for limiting the calculus to $\wfe$-types.

Originally, subtyping was not regular with respect to $\wfe$. Roughly,
all the $\wfe$ preconditions in subtyping were dropped. In this
broader calculus, subtyping transivitiy doesn't hold, because of the
rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.

The problem is deep, as attested by this elaborate counterexample
that is not so easily patched, and directly leads to a counterexample
to preservation.

Consider an environment where $u$ is bound to:
\begin{align*}
\Top & \mlrefine u {\\
&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
}&
\end{align*}

Now, consider the types $S$, $T$, $U$ defined in terms of $u$:
\begin{align*}
S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
T &= u.\mathit{Lower}\\
U &= u.X \refine z {\Ldecl L \Bot \Top}
\end{align*}
We have $S \sub T$ and $T \sub U$, but we cannot derive $S \sub U$ because
$S$ doesn't expand.

Note that $u$ is realizable, since each lower bound is a subtype of
its upper bound. So it is straightforward to turn this counterexample
to subtyping transivity into a counterexample to preservation:
\begin{align*}
&\mlnew u \ldots {
\ \mlapp{\abs x \Top x}{
\ \ \mlapp{\abs f {S \tfun U} f}{
\ \ \ \mlapp{\abs f {S \tfun T} f}{
\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
\ \ \  \ \ \abs x S x}}}}}
\end{align*}

The idea is to start with a function from $S \tfun S$ and cast it
successively to $S \tfun T$ then $S \tfun U$. To typecheck the
expression initially, we need to check $S \sub T$ and $T \sub
U$. After some reduction steps, the first few casts vanish, and the
reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
we need to check $S \sub U$. Thus, we need subtyping transitivity: $S
\sub T$ and $T \sub U$ implies $S \sub U$.
